Convergence Studies of Mass Transport in Disks with Gravitational Instabilities. \\ I. The Constant Cooling Time Case

### 2011-12-06 Paper_1.tex
Changed figure. Added Dick's edits and a second response to referee.
	modified:   Paper_1.tex
	new file:   Response_to_Referee2.txt
	modified:   figures/Figure5.eps

% \slugcomment{Not to appear in Nonlearned J., 45.}

@@

%
\begin{document}

@@

We conduct a convergence study of a protostellar disk, subject to a constant 
global cooling time and susceptible to gravitational instabilities (GIs), at
a time when heating and cooling are roughly balanced. Our goal is to 
determine the gravitational torques produced by GIs, the level to which transport can
be represented by a simple alpha-disk formulation, and to examine 
fragmentation criteria. Four simulations are conducted, identical except for the number
of azimuthal computational grid points used. A Fourier decomposition of 
nonaxisymmetric density structures in  $\cos(m\phi)$, $\sin(m\phi)$ is performed to evaluate 
the amplitudes $A_m$ of these structures. The $A_m$, gravitational torques, and
the effective \citet{shakura1973} $\alpha$  arising from gravitational stresses
are determined for each resolution. We find nonzero $A_m$ for all $m$-values and
that $A_m$  summed over all $m$ is essentially independent of resolution. Because
the number of measurable $m$-values is limited to half the number of azimuthal
grid points, higher-resolution simulations have a larger fraction of their total
amplitude in higher-order structures. These structures act more locally than
lower-order structures. Therefore, as the resolution increases the total 
gravitational stress decreases as well, leading higher-resolution simulations to experience
weaker average gravitational torques than lower-resolution simulations. The 
effective $\alpha$ also depends upon the magnitude of the stresses, thus
$\alpha_{eff}$ also decreases with increasing resolution. Our converged
$\alpha_{eff}$ is consistent with predictions from
an analytic local theory for thin disks by Gammie (2001), but only over many
dynamic times when averaged over a substantial volume of the disk.

@@

In fact, several studies support the idea that angular momentum transport by GIs in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.,][]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006,kratter2010}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in rough balance, 
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, as if it were a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, multiple studies suggest that
an effective $\alpha$-viscosity may be consistent
with transport in GI-active disks for a disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}. 
We note, however, that the these studies all assumed
a {\it local cooling time}, $t_{cool}\Omega =$ constant, 
where $t_{cool}$ is defined by the rate at which cooling decreases the disk's internal energy density
$\epsilon$ by $d\epsilon/dt = \epsilon / t_{cool}$.  In contrast, we and our collaborators use either a {\it global cooling time},
$t_{cool} = $ constant \citep[e.g.,][this paper]{pickett2003, mejia2005}, or use realistic opacities \citep[e.g.,][]{mejiaphd2004, cai2006, cai2008, boley2006, boley2007b,  boley2009} which lead to something in between $t_{cool} = $ constant  and
$t_{cool}\Omega =$ constant. The review by \citet{durisen2007} discusses some consequences of these various
choices. There is a tendency for $t_{cool}\Omega =$ constant to bias a disk toward behavior like that of a
steady-state $\alpha$-disk.

\citet{gammie2001} derived an effective $\alpha$ for the case of a thin gravitationally unstable disk where the heating
caused by GIs was balanced by the local cooling rate \citep[see also][]{pringle1981}.  
Assuming a local cooling time of $t_{cool}\Omega =$ constant,  Gammie found
\begin{equation}
\alpha_{\rm{eff}} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1},
\label{eq:gammie}
\end{equation}
where $\gamma$ is the two-dimensional adiabatic index. 
The two-dimensional index relates
to the three-dimensional adiabatic index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived a similar result when considering three-dimensional protoplanetary disks.
Fully 3D simulations \citep[e.g.,][]{lodato2004, cossins2009} suggest that, at least in the case of imposed 
$t_{cool}\Omega = $ constant cooling, $\alpha_{\rm{\rm{eff}}}$ roughly converges to the value given by equation (\ref{eq:gammie}).

@@

 We assume an equation of state $P = (\gamma -1)\epsilon$, where $P$ is the pressure, $\epsilon$ is the internal energy
 density, and $\gamma$ is the ratio of specific heats, given here by $\gamma = 5/3$. As discussed in \citet{boley2007a},
 the rotation states of $H_2$ are probably not excited in the cold outer parts of the disk ($> 10$ AU); so $\gamma =
 5/3$ is an appropriate choice. The model disks are cooled by decreasing their internal energy according to the prescription $d\epsilon/dt = \epsilon / t_{cool}$, where $t_{cool}$ is the global cooling time, which is set to 2 ORP everywhere.

@@

Figure \ref{fig:Final_Q} shows the Toomre $Q$, as a function of radius, for all four simulations at $t \sim$ 18 ORP.  
%Table \ref{tbl:ams} 
Table 1 lists $Q_{avg}$, the arithmetic mean of $Q$ for each of the asymptotic disks, time-averaged from 12 -- 18 ORP and spatially averaged over 10 -- 40 AU.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject to GIs over this full range.

@@

%Gammie found
%\begin{equation}
%\alpha_{\rm{eff}} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1}
%\end{equation}
%where $\gamma$ is the two dimensional adiabatic index and $t_{cool}$ is a cooling parameter. 
%The two dimensional index relates
%to the three dimensional index, $\Gamma$, by 
%$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
%for a non-self-gravitating
%disk and
%$\gamma = 3 - 2/\Gamma $
%for a strongly self gravitating disk.

Numerical uncertainties render Reynolds stresses difficult to determine accurately. 
This difficulty is mitigated by the fact that 
previous studies have found Reynolds stresses to be small relative to gravitational stresses (see, 
for example, figure 5 of Lodato \& Rice 2004 and figure 10 of Boley et al. 2006).  This  suggests that the Reynolds stress
can be neglected in the evaluation of $\alpha_{\rm{eff}}$.   For this reason, 
here we omit
$T_{\varpi\phi}^{Reyn}$ from the calculation of $\alpha_{\rm{eff}}$. 
We note that \cite{gammie2001}  found
Reynolds and gravitational stresses contributed equally to $\alpha_{\rm{eff}}$ in his study.  However,
his was a local, shearing-sheet, 2D study.  As such, it was unable to capture global effects, such
as those seen in the cited 3D studies.

The gravitational stresses of equation \eqref{alpha1} can be evaluated using the gravitational torque of equation \eqref{torque1}, i.e.,
\begin{equation}
\label{newtonstress}
\begin{split}
 T_{\varpi\phi}^{grav}\left( \varpi \right) & = \frac{1}{2\pi\varpi^2}C_Z\\
                                & = - \frac{1}{2\pi\varpi^2} \int \rho\frac{\partial\Phi}{\partial\phi} {\rm d}V.
\end{split}
\end{equation}
Figure \ref{fig:alpha_v_radius} shows $\alpha_{\rm{eff}}$, time-averaged from 12 to 18 ORP, calculated using equations \eqref{alpha1} and \eqref{newtonstress},
for the four grid resolutions.  For comparison, 
$\alpha_{\rm{eff}}$ predicted by equation \eqref{eq:gammie} from \citet{gammie2001} are plotted for disks.  The lower  Gammie curve applies for the strongly self-gravitating limit while the upper curve 
represents the non-self-gravitating limit.   The $l_{max} = 512$ curve is consistent with these two limits.  However, 
the  $l_{max} = 512$ curve does not represent a converged model.
To determine whether the converged value of $\alpha_{\rm{eff}}$ is consistent with predictions, we examine 
the spatially (10 -- 40 AU) and temporally (12 -- 18 ORP) averaged  $\alpha_{avg}$,
provided in table 1, for each value of  $l_{max}$. A Richardson extrapolation applied to these values  
converges to $\alpha_{avg} = 0.017$ (over the same radial  and time range)
as the grid spacing goes to zero.   This falls
between the $\alpha_{avg}$ values for the Gammie curves 0.014 (strongly self-gravitating) and 0.027 (non-self-gravitating).

@@

\subsection{Convergence and $l_{max}$}

The results above permit an assessment of appropriate azimuthal grids for the problem under consideration and
similar problems. The $l_{max} = 512$ simulation yields a value for $\alpha_{avg}$ of 0.020. This
overestimates the extrapolated converged value of $\alpha_{avg} = 0.017$ by $\sim 18$\%. The $l_{max} = 512$
simulation also overestimates the fractional power in the high-order ($m = 2$ -- 7) modes by essentially the
same percentage. However, it should be noted that the fractional difference between successive doubling of the
resolution has decreased for $l_{max} = 512$. From $l_{max} =64$ to 128 the fractional difference in
$\alpha_{avg}$ is 36\%. This decreases to 26\% for the jump from $l_{max} = 256$ to 512. Furthermore, the
converged value represents infinite azimuthal resolution, which is, of course, impossible to achieve in a
fixed grid scheme.
%It is unclear as to what variation of $\alpha$ between two values of $l_{max}$ would represent convergence as $\alpha$ 
%itself is a dynamically varying quantity, see figure \ref{fig:alphavar}.

%DICK: I made changes here.
In addition to the convergence of $\alpha_{avg}$, table 1 illustrates the convergence of the measured net
inward mass transport rates $\dot{M}$ over the time interval 12 -- 18 ORP. We tabulate the value of
$|\dot{M}|$ spatially averaged over 10 -- 40 AU. The absolute value of the net $\dot{M}$ at each $r$ is taken
prior to the spatial average because $\dot{M}$ varies in sign with $r$. It is the magnitude of the typical net
mass transport rates that we wish to compare. The $|\dot{M}|_{avg}$ measurements show convergence similar to
that of the $\alpha_{avg}$ measurement based on the gravitational torques. This gives us confidence that no
important contributions from Reynolds stresses are being overlooked. Moreover, we have used eq. (22) of
\citet{boley2006} to estimate mass inflow rates based on the measured gravitational torques and find that
these torques are indeed sufficient to account for the measured $\dot{M}$.

@@

\subsection{Fragmentation and Effective $\alpha$s}

@@

Under certain conditions, GIs fail to self-regulate, allowing density perturbations to grow until clumps form from spiral structure (see Durisen et al. 2007 for references).  Under isothermal conditions, fragmentation becomes very likely whenever $Q$ drops below 1.4 
\citep{tomley1991, tomley1994, nelson1998, mayer2002}. When the effective adiabatic index is stiff due to inefficient
cooling, shocks and mass transport can militate against fragmentation, even in disks with initially low $Q$ values
\citep[e.g., ][]{boley2008}. To understand the degree of cooling required to effect clump formation in non-isothermal
disks, \cite{gammie2001} studied the nonlinear behavior of GIs using a series of shearing sheet simulations.  He showed
that  a gravitationally unstable disk is likely to fragment whenever the cooling time is less than half the local
orbital period ($t_{cool}\Omega = 3$).  This work was followed by Rice et al.~(2005), who used global 3D SPH simulations and found that the fragmentation limit was dependent on the adiabatic index.   For disks with $\gamma=5/3$, they found fragmentation 
occurred whenever the local cooling time was less than the local orbital period. For $\gamma=7/5$, the limit was twice the local orbital period.  \cite{rice2005} proposed that this fragmentation threshold was due to a maximum stress that can  be exerted on a disk by gravitational torques.  Using the $\alpha_{\rm{eff}}$ as defined in equation (8), they found that fragmentation could occur whenever $\alpha_{\rm eff}\gtrsim 0.06$ (the dotted line in figure 6).  The difference between the cooling time threshold found by Gammie and the ones found by Rice et al.~are not obviously at odds, as Gammie's study was conducted in 2D and used a 2D adiabatic index of 2.

Recently, these results have been called into question by \cite{meru2011a}, who suggest that our understanding of the
fragmentation criterion is based on simulations that have yet to converge.  In particular, they find that fragmentation
becomes increasingly likely for longer cooling times when the resolution in their 3D SPH simulations is increased,
giving weight to the claims by, e.g., \cite{boss2009}, that fragmentation is a robust result of disk instability. The
Meru \& Bate study uses a prescribed cooling such that $t_{\rm cool}=\beta\Omega^{-1}$, where $\Omega$ is the local
orbital frequency and $\beta$ is some global constant.  This same cooling prescription was used in the Gammie and Rice
et al.~studies.  The reason for the differences is unclear, and may not be simply a resolution effect (see Lodato \&
Clarke 2011, arXiv:1101.2448).  For example, the \cite{gammie2001} have revealed the most stringent cooling criterion of
all the $t_{cool}$ fragmentation studies, and, due to the local approximation Gammie used, these simulations are also at
very high resolution. Moreover, \cite{mejia2005}, who used grid-based simulations with the same code we use here, and  \cite{boley2008}, who used radiative hydrodynamics simulations, also found results that are reasonably consistent with those of Gammie, Rice et al., and \cite{cossins2010}.

@@

Although not a goal of this work, our simulations permit
comments on studies on fragmentation and clump formation.
Most recent studies indicate that our disk should not fragment \citep[e.g., ][]{gammie2001, rice2005, mejia2005, boley2008}.   However, some recent studies \citep[e.g., ][]{boss2009, meru2011a}
%Meru & Bate (2011), 
suggest that fragmentation may be a common result of gravitational instabilities in disks.
We find that as the resolution increases, our disk approaches a self-regulating state that can be reasonably well-described, in the time- and space-average sense, by an $\alpha$-disk parameterization. 
The average Toomre $Q$ increases with increasing resolution,
suggesting that the disk is becoming less susceptible to clump formation with increasing resolution,
consistent with the observation that
the disk simulations do not show a trend toward denser spiral structure as the resolution is increased.
Overall, the disks simulated here are converging toward a state of self-regulation as the resolution is increased, and do not show signs of becoming more susceptible to fragmentation.

@@

SM would like to thank NASA for the generous funding of the
NASA Earth and Space Science Fellowship NNX07AU82H. This
research was also supported in part by NASA Origins of Solar
Systems grants NNG05GN11G and NNX08AK36G. The simulations
and analysis herein were conducted on hardware generously
provided by the Indiana University Information Technology Services
and was supported by the National Science Foundation under
Grant no. ACI-0338618l, OCI-0451237, OCI-0535258, CNS-
0521433 and OCI-0504075. This work was also supported in part
by Shared University Research grants from IBM, Inc., to the Indiana
University.

@@

%DICK: I made changes here.
\begin{table}
\begin{center}
\begin{tabular}{cccccc}
\multicolumn{5}{c}{$Q$, Fourier Amplitudes, $\alpha$, and $\dot{M}$}\\ \hline\hline
$l_{max}$ & $Q_{avg}$   & $\langle A_{tot} \rangle$   & $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ &
$\alpha_{avg}$ & $|\dot{M}|_{avg} ~\mathrm{M_\odot/yr}$\\\hline
64   & 1.26 & 1.96 &  0.76 & 0.069 & 8.9$ \times 10^{-7}$ \\
128 & 1.24 & 1.91 &  0.62 & 0.044 & 3.7$ \times 10^{-7}$ \\
256 & 1.39 & 1.83 &  0.54 & 0.026 & 2.5$ \times 10^{-7}$ \\
512 & 1.48 & 1.95 &  0.49 & 0.020 & 2.4$ \times 10^{-7}$ \\ 
$\infty$ & & &0.43 & 0.017 & \\ \hline
\label{tbl:ams}
\end{tabular}
\caption{ All quantities are time-averaged over 12 -- 18 ORP.
$Q_{avg}$, $\alpha_{avg}$, and $|\dot{M}|_{avg}$  are also spatially
averaged over 10 -- 40 AU.  The entries for ``$\infty$'' are based on Richardson extrapolation,
as explained in the text.}  
\end{center}
\end{table}

>>>>

@@

\begin{document}

@@

We conduct a convergence study of a protostellar disk, subject to a constant 
global cooling time and susceptible to gravitational instabilities (GIs), at
a time when heating and cooling are roughly balanced. Our goal is to 
determine the gravitational torques produced by GIs, the level to which transport can
be represented by a simple alpha-disk formulation, and to examine 
fragmentation criteria. Four simulations are conducted, identical except for the number
of azimuthal computational grid points used. A Fourier decomposition of 
nonaxisymmetric density structures in  $\cos(m\phi)$, $\sin(m\phi)$ is performed to evaluate 
the amplitudes $A_m$ of these structures. The $A_m$, gravitational torques, and
the effective \citet{shakura1973} $\alpha$  arising from gravitational stresses
are determined for each resolution. We find nonzero $A_m$ for all $m$-values and
that $A_m$  summed over all $m$ is essentially independent of resolution. Because
the number of measurable $m$-values is limited to half the number of azimuthal
grid points, higher-resolution simulations have a larger fraction of their total
amplitude in higher-order structures. These structures act more locally than
lower-order structures. Therefore, as the resolution increases the total 
gravitational stress decreases as well, leading higher-resolution simulations to experience
weaker average gravitational torques than lower-resolution simulations. The 
effective $\alpha$ also depends upon the magnitude of the stresses, thus
$\alpha_{\rm eff}$ also decreases with increasing resolution. Our converged
$\alpha_{\rm eff}$ is consistent with predictions from
an analytic local theory for thin disks by Gammie (2001), but only over many
dynamic times when averaged over a substantial volume of the disk.

@@

In fact, several studies support the idea that angular momentum transport by GIs in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.,][]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006,kratter2010,walch2010}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in rough balance, 
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, as if it were a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, multiple studies suggest that
an effective $\alpha$-viscosity may be consistent
with transport in GI-active disks for a disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}. 
We note, however, that the these studies all assumed
a {\it local cooling time}, $t_{\rm cool}\Omega =$ constant, 
where $t_{\rm cool}$ is defined by the rate at which cooling decreases the disk's internal energy density
$\epsilon$ by $d\epsilon/dt = \epsilon / t_{\rm cool}$.  In contrast, we and our collaborators use either a {\it global cooling time},
$t_{\rm cool} = $ constant \citep[e.g.,][this paper]{pickett2003, mejia2005}, or use realistic opacities \citep[e.g.,][]{mejiaphd2004, cai2006, cai2008, boley2006, boley2007b,  boley2009} which lead to something in between $t_{\rm cool} = $ constant  and
$t_{\rm cool}\Omega =$ constant. The review by \citet{durisen2007} discusses some consequences of these various
choices. There is a tendency for $t_{\rm cool}\Omega =$ constant to bias a disk toward behavior like that of a
steady-state $\alpha$-disk.

\citet{gammie2001} derived an effective $\alpha$ for the case of a thin gravitationally unstable disk where the heating
caused by GIs was balanced by the local cooling rate \citep[see also][]{pringle1981}.  
Assuming a local cooling time of $t_{\rm cool}\Omega =$ constant,  Gammie found
\begin{equation}
\alpha_{\rm{eff}} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{\rm cool}]^{-1},
\label{eq:gammie}
\end{equation}
where $\gamma$ is the two-dimensional adiabatic index. 
The two-dimensional index relates
to the three-dimensional adiabatic index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived a similar result when considering three-dimensional protoplanetary disks.
Fully 3D simulations \citep[e.g.,][]{lodato2004, cossins2009} suggest that, at least in the case of imposed 
$t_{\rm cool}\Omega = $ constant cooling, $\alpha_{\rm{eff}}$ roughly converges to the value given by equation (\ref{eq:gammie}).

@@

 We assume an equation of state $P = (\gamma -1)\epsilon$, where $P$ is the pressure, $\epsilon$ is the internal energy
 density, and $\gamma$ is the ratio of specific heats, given here by $\gamma = 5/3$. As discussed in \citet{boley2007a},
 the rotation states of $H_2$ are probably not excited in the cold outer parts of the disk ($> 10$ AU); so $\gamma =
 5/3$ is an appropriate choice. The model disks are cooled by decreasing their internal energy according to the prescription $d\epsilon/dt = \epsilon / t_{\rm cool}$, where $t_{\rm cool}$ is the global cooling time, which is set to 2 ORP everywhere.

@@

Figure \ref{fig:Final_Q} shows the Toomre $Q$, as a function of radius, for all four simulations at $t \sim$ 18 ORP.  
%Table \ref{tbl:ams} 
Table 1 lists $Q_{\rm avg}$, the arithmetic mean of $Q$ for each of the asymptotic disks, time-averaged from 12 -- 18 ORP and spatially averaged over 10 -- 40 AU.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject to GIs over this full range.

@@

%Gammie found
%\begin{equation}
%\alpha_{\rm{eff}} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{\rm cool}]^{-1}
%\end{equation}
%where $\gamma$ is the two dimensional adiabatic index and $t_{\rm cool}$ is a cooling parameter. 
%The two dimensional index relates
%to the three dimensional index, $\Gamma$, by 
%$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
%for a non-self-gravitating
%disk and
%$\gamma = 3 - 2/\Gamma $
%for a strongly self gravitating disk.

As discussed in section \ref{sec:converge}, numerical uncertainties render Reynolds stresses difficult to
determine accurately.  This difficulty is mitigated somewhat by the fact that previous global three
dimensional studies have found Reynolds stresses to be small relative to gravitational stresses (see, for
example, figure 5 of Lodato \& Rice 2004 and figure 10 of Boley et al. 2006).  Here we omit
$T_{\varpi\phi}^{Reyn}$ from the calculation of $\alpha_{\rm{eff}}$ and show in section \ref{sec:converge}
that, in fact, the gravitational torques are sufficient to account for all the mass transport.  We note that
\cite{gammie2001} found Reynolds and gravitational stresses contributed equally to $\alpha_{\rm{eff}}$ in his
study.  However, his was a local, shearing-sheet, 2D study.  As such, it was unable to capture global effects,
such as those seen in the cited 3D studies.

The gravitational stresses of equation \eqref{alpha1} can be evaluated using the gravitational torque of equation \eqref{torque1}, i.e.,
\begin{equation}
\label{newtonstress}
\begin{split}
 T_{\varpi\phi}^{grav}\left( \varpi \right) & = \frac{1}{2\pi\varpi^2}C_Z\\
                                & = - \frac{1}{2\pi\varpi^2} \int \rho\frac{\partial\Phi}{\partial\phi} {\rm d}V.
\end{split}
\end{equation}
Figure \ref{fig:alpha_v_radius} shows $\alpha_{\rm{eff}}$, time-averaged from 12 to 18 ORP, calculated using equations \eqref{alpha1} and \eqref{newtonstress},
for the four grid resolutions.  For comparison, 
$\alpha_{\rm{eff}}$ predicted by equation \eqref{eq:gammie} from \citet{gammie2001} are plotted for disks.  The lower  Gammie curve applies for the strongly self-gravitating limit while the upper curve 
represents the non-self-gravitating limit.   The $l_{max} = 512$ curve is consistent with these two limits.  However, 
the  $l_{max} = 512$ curve does not represent a converged model.
To determine whether the converged value of $\alpha_{\rm{eff}}$ is consistent with predictions, we examine 
the spatially (10 -- 40 AU) and temporally (12 -- 18 ORP) averaged  $\alpha_{\rm avg}$,
provided in table 1, for each value of  $l_{max}$. A Richardson extrapolation applied to these values  
converges to $\alpha_{\rm avg} = 0.017$ (over the same radial  and time range)
as the grid spacing goes to zero.   This falls
between the $\alpha_{\rm avg}$ values for the Gammie curves 0.014 (strongly self-gravitating) and 0.027 (non-self-gravitating).

@@

\subsection{Convergence and $l_{max}$}\label{sec:converge}

The results above permit an assessment of appropriate azimuthal grids for the problem under consideration and
similar problems. The $l_{max} = 512$ simulation yields a value for $\alpha_{\rm avg}$ of 0.020. This
overestimates the extrapolated converged value of $\alpha_{\rm avg} = 0.017$ by $\sim 18$\%. The $l_{max} = 512$
simulation also overestimates the fractional power in the high-order ($m = 2$ -- 7) modes by essentially the
same percentage. However, it should be noted that the fractional difference between successive doubling of the
resolution has decreased for $l_{max} = 512$. From $l_{max} =64$ to 128 the fractional difference in
$\alpha_{\rm avg}$ is 36\%. This decreases to 26\% for the jump from $l_{max} = 256$ to 512. Furthermore, the
converged value represents infinite azimuthal resolution, which is, of course, impossible to achieve in a
fixed grid scheme.
%It is unclear as to what variation of $\alpha$ between two values of $l_{max}$ would represent convergence as $\alpha$ 
%itself is a dynamically varying quantity, see figure \ref{fig:alphavar}.

%DICK: I made changes here.
In addition to the convergence of $\alpha_{\rm avg}$, table 1 illustrates the convergence of the measured net
inward mass transport rates $\dot{M}$ over the time interval 12 -- 18 ORP. We tabulate the value of
$|\dot{M}|$ spatially averaged over 10 -- 40 AU. The absolute value of the net $\dot{M}$ at each $r$ is taken
prior to the spatial average because $\dot{M}$ varies in sign with $r$. It is the magnitude of the typical net
mass transport rates that we wish to compare. The $|\dot{M}|_{\rm avg}$ measurements show convergence similar to
that of the $\alpha_{\rm avg}$ measurement based on the gravitational torques. This gives us confidence that no
important contributions from Reynolds stresses are being overlooked. Moreover, we have used eq. (22) of
\citet{boley2006} to estimate mass inflow rates based on the measured gravitational torques and find that
these torques are indeed sufficient to account for the measured $\dot{M}$.

The principal reasons for not including the contributions from the Reynolds stress are twofold:

1) As
elucidated in the above paragraph, the gravitational torques are sufficient to account for the measured
$\alpha$ and $\dot{M}$, and all three converge as resolution is increased. There does not appear to be an
additional source of stress required to explain the observed mass transport.

2) It is difficult to measure the Reynolds stress accurately in a nonlocal, strongly nonlinear calculation. The difficulty arises from the fact that the Reynolds
stress is typically defined as 
\begin{equation}
\label{eq:reynstress}
R_{ij} = \rho u'_i u'_j
\end{equation}
where $u'_i$ and $u'_j$ are the fluctuations in the velocity field. The fluctuating pieces of the velocity
field are defined as $u'_i = u_i - u_i({\rm avg})$, but in a gravitationally unstable disk with extended
non-axisymmetric spiral structures, the proper definition of the ``mean'' flow, $u_i({\rm avg})$, is unclear.

In attempting to measure the Reynolds stresses, we tried using several different
definitions for $u_{\phi}({\rm avg})$ and obtained results that varied dramatically. We used several local
averaging schemes as well as an azimuthal average. Although Reynolds stresses are presented for other global
simulations, \citep[e.g.][]{lodato2004}, the precise methods used to determine
the stress values are not always elucidated. Because of this, it is difficult to make a direct comparison.
The particulars of the method chosen can have a significant effect on the computed values of the
Reynolds stress.

@@

\subsection{Fragmentation and Effective Alphas}

@@

Under certain conditions, GIs fail to self-regulate, allowing density perturbations to grow until clumps form from spiral structure (see Durisen et al. 2007 for references).  Under isothermal conditions, fragmentation becomes very likely whenever $Q$ drops below 1.4 
\citep{tomley1991, tomley1994, nelson1998, mayer2002}. When the effective adiabatic index is stiff due to inefficient
cooling, shocks and mass transport can militate against fragmentation, even in disks with initially low $Q$ values
\citep[e.g., ][]{boley2008}. To understand the degree of cooling required to effect clump formation in non-isothermal
disks, \cite{gammie2001} studied the nonlinear behavior of GIs using a series of shearing sheet simulations.  He showed
that  a gravitationally unstable disk is likely to fragment whenever the cooling time is less than half the local
orbital period ($t_{\rm cool}\Omega = 3$).  This work was followed by Rice et al.~(2005), who used global 3D SPH simulations and found that the fragmentation limit was dependent on the adiabatic index.   For disks with $\gamma=5/3$, they found fragmentation 
occurred whenever the local cooling time was less than the local orbital period. For $\gamma=7/5$, the limit was twice the local orbital period.  \cite{rice2005} proposed that this fragmentation threshold was due to a maximum stress that can  be exerted on a disk by gravitational torques.  Using the $\alpha_{\rm{eff}}$ as defined in equation (8), they found that fragmentation could occur whenever $\alpha_{\rm eff}\gtrsim 0.06$ (the dotted line in figure 6).  The difference between the cooling time threshold found by Gammie and the ones found by Rice et al.~are not obviously at odds, as Gammie's study was conducted in 2D and used a 2D adiabatic index of 2.

Recently, these results have been called into question by \cite{meru2011a}, who suggest that our understanding of the
fragmentation criterion is based on simulations that have yet to converge.  In particular, they find that fragmentation
becomes increasingly likely for longer cooling times when the resolution in their 3D SPH simulations is increased,
giving weight to the claims by, e.g., \cite{boss2009}, that fragmentation is a robust result of disk instability. The
Meru \& Bate study uses a prescribed cooling such that $t_{\rm cool}=\beta\Omega^{-1}$, where $\Omega$ is the local
orbital frequency and $\beta$ is some global constant.  This same cooling prescription was used in the Gammie and Rice
et al.~studies.  The reason for the differences is unclear, and may not be simply a resolution effect \citep{lodato2011}.  For example, the \citeauthor{gammie2001} work revealed the most stringent cooling criterion of
all the $t_{\rm cool}$ fragmentation studies, and, due to the local approximation Gammie used, these simulations are also at
very high resolution. Moreover, \cite{mejia2005}, who used grid-based simulations with the same code we use
here, and  \cite{boley2008}, who used radiative hydrodynamics simulations, also found results that are
reasonably consistent with those of \citet{gammie2001}, \citet{rice2005}, and \cite{cossins2010}. Very recent
work by \citet{paardekooper2011} supporting the \citeauthor{gammie2001} fragmentation criterion suggests that
convergence results for fragmentation may be sensitive to the presence of sharp edges and the manner in which
fragmentation is approached.

@@

Although not a goal of this work, our simulations can be used to comment on studies of fragmentation and clump
formation. Previous work indicates that our disk should not fragment due to the long cooling times that we
use \citep[e.g.,][]{gammie2001, rice2005, mejia2005, boley2008}. However, some recent studies \citep[e.g.,][]{boss2009,meru2011a} suggest that previously determined fragmentation criteria too restrictive due to poor
resolution. We find here that as the resolution is increased, our disk approaches a self-regulating state that
can be reasonably well-described, in the time- and space-average sense, by an $\alpha$-disk
parameterization. The average Toomre $Q$ also increases with increasing resolution, suggesting that the disk
becomes less susceptible to clump formation as the resolution is increased.  This is consistent with our
simulations not showing a trend toward denser spiral structure with increasing resolution. Moreover, recent
work by \citet{paardekooper2011} showed that if a high cooling rate is approached slowly, the
\citet{gammie2001} cooling criterion is recovered for all resolutions considered.  We do note that they find
non-convergence with resolution when sharp edges in disks are present, but as the authors suggest, this may be
numerically driven (see also \citet{lodato2011} on numerical convergences issues in regards to fragmentation). Overall, the disks simulated here are converging toward a state of self-regulation as the
resolution is increased, and do not show signs of becoming more susceptible to fragmentation.

@@

We would like to thank the anonymous referee for very useful comments that substantially improved this manuscript. SM
would like to thank NASA for the generous funding of the NASA Earth and Space Science Fellowship
NNX07AU82H. This research was also supported in part by NASA Origins of Solar Systems grants NNG05GN11G and
NNX08AK36G. The simulations and analysis herein were conducted on hardware generously provided by the Indiana
University Information Technology Services and was supported by the National Science Foundation under Grant
no. ACI-0338618l, OCI-0451237, OCI-0535258, CNS- 0521433 and OCI-0504075. This work was also supported in part
by Shared University Research grants from IBM, Inc., to the Indiana University. This work used the Extreme Science and Engineering Discovery Environment (XSEDE), which is supported by National Science Foundation grant number OCI-1053575

@@

%DICK: I made changes here.
\begin{table}
\begin{center}
\begin{tabular}{cccccc}
\multicolumn{5}{c}{$Q$, Fourier Amplitudes, $\alpha$, and $\dot{M}$}\\ \hline\hline
$l_{max}$ & $Q_{\rm avg}$   & $\langle A_{tot} \rangle$   & $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ &
$\alpha_{\rm avg}$ & $|\dot{M}|_{\rm avg} ~\mathrm{M_\odot/yr}$\\\hline
64   & 1.26 & 1.96 &  0.76 & 0.069 & 8.9$ \times 10^{-7}$ \\
128 & 1.24 & 1.91 &  0.62 & 0.044 & 3.7$ \times 10^{-7}$ \\
256 & 1.39 & 1.83 &  0.54 & 0.026 & 2.5$ \times 10^{-7}$ \\
512 & 1.48 & 1.95 &  0.49 & 0.020 & 2.4$ \times 10^{-7}$ \\ 
$\infty$ & & &0.43 & 0.017 & \\ \hline
\label{tbl:ams}
\end{tabular}
\caption{ All quantities are time-averaged over 12 -- 18 ORP.
$Q_{\rm avg}$, $\alpha_{\rm avg}$, and $|\dot{M}|_{\rm avg}$  are also spatially
averaged over 10 -- 40 AU.  The entries for ``$\infty$'' are based on Richardson extrapolation,
as explained in the text.}  
\end{center}
\end{table}
